---
title: Testing Linear Hypotheses
subtitle: Lecture 4
format: 
  html: 
    output-file: lecture-4.html
    other-links:
      - text: "PDF with Stata"
        href: "lecture-4-notes-Stata.pdf"
        icon: file-pdf
      - text: "PDF with R"
        href: "lecture-4-notes-R.pdf"
        icon: file-pdf
  revealjs: 
    output-file: lecture-4-slides.html
    footer: "[Lecture 4](lecture-4.qmd) - St Andrews Business School - Martinmas 2026/27"
engine: knitr
---

```{r}
#| include: false
#| cache: false

library(Statamarkdown)
knitr::opts_chunk$set(collectcode = TRUE)
```


```{r}
#| include: false

# Packages
library(haven)    
library(ggplot2)

source(here::here("config.R"))
```

```{stata}
*| include: false

do "config.do"
```

::: {.content-visible unless-format="revealjs"}
## Reading

This lecture covers material from @wooldridge2025:

- Chapter 4.4 - "Testing Hypotheses about a Single Linear Combination of the Parameters"
- Chapter 4.5 - "Testing Multiple Linear Restrictions"
- Chapter 7.4 - "Interactions Involving Dummy Variables"

Note: I have edited the syllabus slightly to move the topic of heteroskedasticity to Week 5 with asymptotics.
:::

:::: {.content-visible when-format="revealjs"}
## Recap

::: incremental
1. Properties of OLS

   - Bias (Mean) - MLR 1-4
   - Efficiency (Variance) - MLR 1-5
   - Distribution - MLR 1-6

2. Simple hypothesis tests
  
   - Critical value, $p$-value, and confidence interval
   - $z$-stat (know error variance)
   - $t$-stat (estimate error variance)
   - Power of the test
:::
 
## Power: example

**Why it's important?**

::: {.fragment .fade-in}
Suppose, you want you evaluate a policy - like the [Scottish Child Payment](https://www.mygov.scot/scottish-child-payment).

  - Weekly payment of £28.20
  - For children under 16
  - One of [5 policies](https://www.mygov.scot/best-start-grant-best-start-foods) designed to reduce child povery in Scotland. 

:::
   
## 

You estimate a model to test whether recipients of the childpayment payment ($CP$) are less likely to be recorded as living in poverty.[^1]

[^1]: This is what we call a linear probability model, as the outcome is a binary variable (see not on [Interpreting Linear Models](material-interpretation.qmd)). 

$$
  poor_i = \alpha + \beta CP_i + \mathbf{x}_i\gamma + \varepsilon_i
$$

In Scotland, the child povery rate is 21% ([source](https://www.gov.scot/news/latest-poverty-statistics-published/))

## 

Your estimate of $\beta$ is,

$$
  -0.02 \;(se = 0.016)
$$

- The $t$-stat is $1.25$ (which is not significant at 10% level for any reasonable sample size)

- If the result was significant it would mean that the CP reduced child poverty by $\approx 10\%$. A sizeable reduction.

- BUT, you cannot reject the null that the SCP has had no effect, $H_0: \beta=0$.

::: {.fragment .fade-in}
Is that because the policy *truly* had no effect or because your test has little **statistical power** to reject the null when the effect is indeed $\beta=-0.02$. 

$$
  Pr(\text{Reject}\;H_0|\beta=-0.02) = ?
$$

- The power of the test ultimately depends on the **variance** of the estimator. 

:::
   
::::

# Hypothesis Tests

## Simple hypotheses

So far, we have considered hypotheses that involve a single estimator:

::: incremental

1. $H_0: \beta_j=\kappa_0$

2. $H_0: \beta_j\geq \kappa_0$

3. $H_0: \beta_j\leq \kappa_0$

:::

## Linear combinations

Today, we consider hypotheses of the form:

$$ 
  H_0: a_0\beta_0 + a_1 \beta_1 + \cdots + a_k \beta_k = \kappa_0 
$$

for constants $\{a_j\}_{j=0,1,\dots,k}$. This is a **linear combination of parameters**.

::: {.fragment .fade-in}
For example, $H_0: \beta_1= 2\beta_2-0.5$ can be written as, 

$$
  H_0: \underbrace{0}_{a_0}\beta_0+\underbrace{1}_{a_1}\beta_1 +\underbrace{(- 2)}_{a_2}\beta_2 = \underbrace{-0.5}_{\kappa_0}
$$

:::

## Multiple linear combinations

Similarly, we could have $q$ of these statements,

$$
  \begin{aligned}
   H_0: a^1_0\beta_0 + a^1_1 \beta_1 + \cdots + a^1_k \beta_k &= \kappa^1_0 \\
    a^2_0\beta_0 + a^2_1 \beta_1 + \cdots + a^2_k \beta_k &= \kappa^2_0 \\
    &\vdots \\
    a^q_0\beta_0 + a^q_1 \beta_1 + \cdots + a^q_k \beta_k &= \kappa^q_0 \\
  \end{aligned}
$$

## Single linear combinations

Consider the following example related to environmental and health economics

$$
  \ln(PM2.5_\ell) = \beta_0 + \beta_1 EV_\ell + \beta_2 hybrid_\ell + \mathbf{x}_\ell\gamma + u_\ell
$$

where, 

::: incremental

- $PM2.5$ is micrograms of particulate matter (with diameter $\leq 2.5$ micrometres) per cubic metre in location $\ell$. 

  - standard measure for health effects as absorbable by lungs and bloodstream
  
- $EV$ and $hybrid$ are number of registered vehicles (/1000) by fuel type in location $\ell$. 

:::

## Example: PM2.5

The following estimates use local authority level data for 2019 in the UK. 

::: {.panel-tabset}

### Stata 

```{stata}
*| echo: true
*| code-fold: true
*| output: true

* Open CSV data
import delimited "$data_dir\pm25.csv", clear

* Estimate univariate linear regression
reg ln_pm25 ev_k hybrid_k petrol_k diesel_k ln_pop
```

### R

```{r}
#| echo: true
#| code-fold: true
#| warning: false
    

# Load the Stata file
pm25_data <- read.csv(file.path(data_dir, "pm25.csv"))

# Estimate univariate linear regression
model1 <- lm(ln_pm25 ~ ev_k + hybrid_k + petrol_k + diesel_k + ln_pop, data = pm25_data)

# View results
summary(model1)
```
:::

## Null hypotheses

From a public policy perspective (if you consider designing subsidies), you want to estimate both $\beta_1$ and $\beta_2$, BUT also know if

$$
  H_0: \beta_1=\beta_2
$$

That is, should the policy discriminate bewteen electrified fuel types?

- Engineers may be able provide mechanical differences, but they can't tell you about differences in usage (driver behaviour) which is important for realized pollution mitigation.

:::{.callout-warning}
Do not take these estimates too seriously. This model is a classic example of selection into "treatment": those who live in higher pollution areas (e.g. dense urban centres) are also more likely to buy EVs. A better research design would use panel data to evaluate this question dyanmically.
:::

## Alternative hypothesis

The alternative in this case, may be 

$$
  H_1: \beta_1\neq \beta_2
$$

But may also be, 

$$
  H_1: \beta_1<\beta_2
$$

or

$$
  H_1: \beta_1>\beta_2
$$

if there is a strong case for ruling out one case [see @wooldridge2025, p. 139].

## Test statistic

For $H_0: \beta_1=\beta_2\Rightarrow H_0:\beta_1-\beta_2=0$


$$
  \text{test-stat} = \frac{\hat{\beta}_1-\hat{\beta}_2}{\sqrt{Var(\hat{\beta}_1-\hat{\beta}_2)}}
$$

:::{.callout-note}
The more general case would be $H_0: a_1\beta_1+a_2\beta_2=\kappa_0$

$$
  \text{test-stat} = \frac{a_1\hat{\beta}_1+a_2\hat{\beta}_2-\kappa_0}{\sqrt{Var(a_1\hat{\beta}_1+a_2\hat{\beta}_2)}}
$$

which can be extended to include more parameters. 
:::

## Variance

The key to understanding and executing this test is,[^2]

[^2]: We drop the $Var(\cdot |X)$ notation from [Lecture 3](lecture-3-multi.qmd). However, all these variances are conditional on the sample values of the independent variables. 

$$
  Var(\hat{\beta}_1-\hat{\beta}_2) \neq Var(\hat{\beta}_1) + Var(\hat{\beta}_2)
$$

::: {.fragment .fade-in}
This is because, 

$$
  \begin{aligned}
  Var(a_1z_1+a_2z_2) =& Var(a_1z_1) + Var(a_2z_2) + 2\times Cov(a_1z_1,a_2z_2) \\
  =&a_1^2Var(z_1) + a_2^2 Var(z_2) + 2\times a_1a_2Cov(z_1,z_2)
  \end{aligned}
$$

for random variables $z_1$ and $z_2$ and constants $a_1$ and $a_2$.
:::

## 

So, 

$$
  Var(\hat{\beta}_1-\hat{\beta}_2) = Var(\hat{\beta}_1) + Var(\hat{\beta}_2) - \textcolor{red}{2\times}\underbrace{\textcolor{blue}{Cov(\hat{\beta}_1,\hat{\beta}_2)}}_{\neq 0}
$$

::: {.fragment .fade-in}
**Why $-2$?**

- For this $H_0$, 

  $$
    \Rightarrow H_0: \underbrace{\beta_1-\beta_2 = 0}_{a_1\beta_1 + a_2\beta_2 = \kappa_0}
  $$
  
  So, $a_1 = 1$ and $a_2 = -1$. 

:::

## Distribution

Under MLR 1-6, the **conditional** distribution of this test-statistic is $t_{n-k-1}$

$$
  t\text{-stat} = \frac{\hat{\beta}_1-\hat{\beta}_2}{\sqrt{Var(\hat{\beta}_1-\hat{\beta}_2)}}
  \underset{H_0}{\sim} t_{n-k-1}
$$

**under the null**.

- You can use the same steps from [Lecture 3](lecture-3-multi.qmd) to conduct inference: $p$-value, critical value. 

- We tend not to construct confidence intervals in such cases. 

- Similarly, you can compute the power under the alternative $\beta_1-\beta_2 = \kappa$. 

## Example: PM2.5

::: {.panel-tabset}

### Stata 

```{stata}
*| echo: true
*| output: true

* t-test
lincom ev_k - hybrid_k
```

### R

In R, you can use the `marginaleffects` package. However, this is NOT available in AppsAnywhere. Do not worry, there is an alternative coming up.

```{r}
#| echo: true
#| warning: false
    
# Load package
library(marginaleffects)

# Estimate linear hypothesis
hypotheses(model1, "ev_k - hybrid_k = 0")
```
:::


## Multiple linear hypotheses

Another common form of hypothesis concerns **multiple** linear combinations of estimators. 

$$
  \begin{aligned}
   H_0: a^1_0\beta_0 + a^1_1 \beta_1 + \cdots + a^1_k \beta_k &= \kappa^1_0 \\
    a^2_0\beta_0 + a^2_1 \beta_1 + \cdots + a^2_k \beta_k &= \kappa^2_0 \\
    &\vdots \\
    a^q_0\beta_0 + a^q_1 \beta_1 + \cdots + a^q_k \beta_k &= \kappa^q_0 \\
  \end{aligned}
$$

::::{.content-visible unless-format="revealjs"}
:::{.callout-tip title="Linear transformation" collapse=true}
You can  think of this as a linear transformation of $\beta$. 

$$
  H_0: R\beta = r
$$

Where $R$ is a $q\times k$ matrix of constants and $r$ a $q$-dimensional column vector of constants. 
:::
::::

## Example: PM2.5

In our example of 

$$
  PM2.5_\ell = \beta_0 + \beta_1 EV_\ell + \beta_2 hybrid_\ell + \mathbf{x}_\ell\gamma + u_\ell
$$

This could be as simple as, 

$$
  H_0: \beta_1 = 0\;\text{and}\; \beta_2 = 0
$$

:::{.callout-caution title="Multiple hypothesis testing"}
Why not test $H^{(1)}_0: \beta_1=0$ and then $H^{(2)}_0: \beta_2 = 0$ as separate individual hypotheses? The combined size of these tests would not be $\alpha$. Why? The estimators are are correlated which implies:

$$
  Pr\big(\text{Reject}\;H^{(1)}_0\;\text{or}\;H^{(2)}_0 \big|\beta_1=0,\beta_1=0\big) \neq \alpha
$$

:::

::::{.content-visible unless-format="revealjs"}
:::{.callout-tip title="Multiple hyopthesis testing" collapse=true}
There are a number of ways to correctly do multiple hypothesis testing. One is the [Bonferroni correction](https://en.wikipedia.org/wiki/Bonferroni_correction) which gets you to reject separate hypotheses sequentially using an algorithim. It does so by ensuring that the family-wise error rate (the probability of rejecting at least one true $H_0$ across a "family" of hypotheses) is $\leq \alpha$. 
:::
::::

## Test

The test is performed by considering two models:

1. **Unrestricted** model; using our example,

$$
  \ln(PM2.5_\ell) = \beta_0 + \beta_1 EV_\ell + \beta_2 hybrid_\ell + \mathbf{x}_\ell\gamma + u_\ell
$$

   - storing the $SSR_{ur}$ from this model

::: {.fragment .fade-in}

2. **Restricted model**, which imposes $H_0: \beta_1 = 0\;\text{and}\; \beta_2 = 0$,

$$
  \ln(PM2.5_\ell) = \beta_0 + \mathbf{x}_\ell\gamma + u_\ell
$$

   - storing the $SSR_{r}$ from this model

:::

## Test statistic

The test static that is then given by, 

$$
  F\text{-stat} = \frac{(SSR_{r}-SSR_{ur})/q}{SSR_{ur}/df_{ur}}
$$

::: {.fragment .fade-in}
where, 

- $SSR_r$ and $SSR_{ur}$ come from the restricted ($r$) and unrestricted models ($ur$).[^3]

[^3]: This inequality is how you can remember the order. The $F$-stat is strictly positive.

  $$
    SSR_{r}-SSR_{ur}>0
  $$
 
:::

## Test statistic

The test static that is then given by, 

$$
  F\text{-stat} = \frac{(SSR_{r}-SSR_{ur})/q}{SSR_{ur}/df_{ur}}
$$

and, 

- $q$ is the number of restrictions (i.e. equations tested) and is given by the difference in residual degrees of freedom

  $$
    q = df_r-df_{ur}
  $$

## Distribution

Under the null, $F$-stat has an $F$ distribution (hence the name)

$$
  F\text{-stat} \underset{H_0}{\sim}F_{q,n-k-1}
$$

We reject $H_0$ if 


$$
  F\text{-stat} > F_{q,n-k-1}^{-1}(1-\alpha) = f_{q,n-k-1,1-\alpha}
$$

where $F_{q,n-k-1}^{-1}(1-\alpha)$ is the $1-\alpha$ percentile of the distribution. 

## Rejection rule

$$
  F\text{-stat} > f_{q,n-k-1,1-\alpha}
$$

::: incremental

- **Always a one-sided test**.[^4]

[^4]: This does not mean that $F$-distributed tests are always one-sided, but rather that multiple linear hypotheses of this form are. 

- Doesn't mean that $H_1$ is one-sided. 

  - You cannot tell which direction the deviation from the null is, because F-stat uses squared-deviations (i.e. sign is lossed)  . 

- $H_1:\; H_0\;\text{false}$, or $H_1: \beta_1\neq0\;\text{or}\; \beta_2\neq0$

:::

::::{.content-visible unless-format="revealjs"}
::: {.callout-tip title="$F$ distribution" collapse="true"}

The $F$ distribution is defined as the ratio of two chi-squared distributions, 

$$
  F_{q,m} = \frac{\chi^2_q/q}{\chi^2_m/m}
$$

each normalized by their corresponding degrees of freedom. 

:::
::::

## Alternative form

Since $SSR = (1-\mathbf{R}^2)SST$ and $SST_{r}=SST_{ur}$, 

We can write the $F$-stat as, 

$$
  F\text{-stat} = \frac{(\mathbf{R}^2_{ur}-\mathbf{R}^2_{r})/q}{(1-\mathbf{R}^2_{ur})/(n-k-1)}
$$

Note, 

- The order switches in the numerator, but remains positive since $\mathbf{R}^2_{ur}>\mathbf{R}^2_{r}$

- $n-k-1=df_{ur}$, with $k$ the number of parameters in the unrestricted model

## Example: PM2.5

::: {.panel-tabset}

### Stata 

```{stata}
*| echo: true
*| output: true

* F-test
test ev_k hybrid_k
```

### R

In R, you can use the `car` package, which is available in AppsAnywhere.

```{r}
#| echo: true
#| warning: false
    
# Load package
library(car)

# Estimate linear hypothesis
linearHypothesis(model1, c("ev_k = 0", "hybrid_k = 0"))
```
:::

## Revisiting single linear restriction

You can think of a single linear hypothesis as a special case of multiple where $q=1$. For example, 

$$
  H_0: \textcolor{red}{\beta_1} = \textcolor{blue}{\beta_2}
$$

- Restricted model:

  $$
    \begin{aligned}
    \ln(PM2.5_\ell) =& \beta_0 + \textcolor{red}{\beta_1} EV_\ell + \underbrace{\textcolor{red}{\beta_1}}_{=\textcolor{blue}{\beta_2}} hybrid_\ell + \mathbf{x}_\ell\gamma + u_\ell \\
    = & \beta_0 + \textcolor{red}{\beta_1} (EV_\ell + hybrid_\ell)   + \mathbf{x}_\ell\gamma + u_\ell
    \end{aligned}            
  $$
  
  A regression with a new transformed variable: $EV_\ell + hybrid_\ell$

## 

Moreover, you can show that for $q=1$

$$
  F\text{-stat} = \frac{SSR_{r}-SSR_{ur}}{SSR_{ur}/(n-k-1)} = (t\text{-stat})^2
$$

## Example: PM2.5

::: {.panel-tabset}

### Stata 

```{stata}
*| echo: true
*| code-fold: true
*| output: true

* t-test
lincom ev_k - hybrid_k

* F-test
test ev_k=hybrid_k

* check
dis sqrt(r(F))
```

### R

```{r}
#| echo: true
#| code-fold: true
#| warning: false
    
# t-test
hypotheses(model1, "ev_k - hybrid_k = 0")

# F-test
linearHypothesis(model1, "ev_k = hybrid_k")
```
:::



## Regression significance

A special case of this form of test concerns the joint significance of all the *slope* parameters in a model:

$$
  H_0: \beta_1 = 0,\; \beta_2 = 0,\;\dots,\;\beta_k=0
$$

In this case, 

- The restricted model is a regression on a constant (alone) where $\Rightarrow \mathbf{R}^2_{r}=0$

- And $\mathbf{R}^2_{ur}$ is simply the $\mathbf{R}^2$ from the model

- The number of restrictions is $q=k$

##

We define the $F$-stat as, 

  $$
    F\text{-stat} = \frac{\mathbf{R}^2/k}{(1-\mathbf{R}^2)/(n-k-1)}
  $$

## Stata output {.scrollable}

This test is reported as standard with the Stata output as: 

- $\texttt{F( , )}$: $F$-stat 
- $\texttt{Prob > F}$: corresponding $p$-value of $F$-test 

```{stata}
*| echo: false
*| output: true

* Estimate linear regression
reg ln_pm25 ev_k hybrid_k petrol_k diesel_k ln_pop
```



# Structural Change

## Dummy variables: recap

Recall, when we have the model:

$$
  y_i = \beta_0 + \beta_1 d_i + u_i
$$

where $d_i\in\{0,1\}$. Then, 

$$
  \beta_1 = E[y_i|d_i=1]-E[y_i|d_i=0]
$$

## Dummy variable interactions

Now, suppose the model was 

$$
  y_i = \beta_0 + \beta_1 x_i + \beta_2 d_i + \beta_3 x_i\times d_i + u_i
$$

where $x_i$ is continuous. 

- What is the interpretation of each parameter?


## 

$$
  y_i = \beta_0 + \beta_1 x_i + \beta_2 d_i + \beta_3 x_i\times d_i + u_i
$$

::: incremental

- $\beta_0=E[y_i|\textcolor{blue}{d_i=0},x_i=0]$

- $\beta_1 = \frac{dE[y_i |\textcolor{blue}{d_i=0},x_i]}{dx_i}$

- $\beta_2=E[y_i|\textcolor{red}{d_i=1},x_i=0]-E[y_i|\textcolor{blue}{d_i=0},x_i=0]$

- $\beta_3 = \frac{dE[y_i |\textcolor{red}{d_i=1},x_i]}{dx_i}-\frac{dE[y_i |\textcolor{blue}{d_i=0},x_i]}{dx_i}$

:::

## 

Another way of thinking about it:

- For $d_i=0$, the model is: 

  $$
    y_i = \textcolor{blue}{\beta_0} + \textcolor{blue}{\beta_1} x_i + u_i
  $$

- For $d_i = 1$, the model is:

  $$
    y_i = \beta_0 + \beta_1 x_i + \beta_2 +  \beta_3 x_i + u_i
  $$

  Which can be rewritten as:
  
  $$
    y_i = \underbrace{(\beta_0 + \beta_2)}_{\textcolor{red}{\gamma_0}} +  \underbrace{(\beta_1 + \beta_3)}_{\textcolor{red}{\gamma_1}} x_i + u_i
  $$

##

We have two models: for $d_i=0$ and $d_i=1$. 

- $\beta_2$ is the difference in intercept

- $\beta_3$ is the difference in slope



## Example

The typical example is the returns to education for men and women [as in @wooldridge2025, p. 243]

- But, given what we know about the 'parent penalty' [@kleven2019child], it seems more pertinent to focus on experience

  $$
    \ln(wage_i) = \beta_0 + \beta_1 exp_i + u_i
  $$

##

::: {.panel-tabset}

### Stata

```{stata}
*| echo: true
*| code-fold: true
*| stata.fig: true
*| fig-align: center
*| out-width: "67%"

use "$data_dir\morg24_bothgen.dta", clear

* Scatter
twoway (scatter lnwage exp, mc(blue%50)) (lfit lnwage exp, lwidth(*2)), by(sex,  legend(off) note("") ) xtitle("Potential Experience (=Age-(Years of Education +6))")

```

### R

```{r}
#| echo: true
#| code-fold: true
#| fig-align: center
#| out-width: "67%"

wage_data <- read_dta(file.path(data_dir, "morg24_bothgen.dta"))

ggplot(wage_data, aes(x = exp, y = lnwage)) +
  geom_point(colour = "blue", alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5) +
  facet_wrap(~ as_factor(sex)) +
  labs(x = "Potential Experience (=Age-(Years of Education +6))") +
  theme_minimal()
```

:::

##

If we overlay the fitted lines we get

::: {.panel-tabset}

### Stata

```{stata}
*| echo: true
*| code-fold: true
*| stata.fig: true
*| fig-align: center
*| out-width: "67%"

* overlay fitted lines
twoway (lfit lnwage exp if sex==1, lc(blue)) (lfit lnwage exp if sex==2, lc(red) lwidth(*2)), xtitle("Potential Experience (=Age-(Years of Education +6))") legend(order(1 "Male" 2 "Female") r(2) ring(0) pos(5)) ylabel(2(1)6)
```

### R

```{r}
#| echo: true
#| code-fold: true
#| fig-align: center
#| out-width: "67%"

wage_data$group <- factor(wage_data$sex, levels = c(1, 2),
                          labels = c("Male", "Female"))

ggplot(wage_data, aes(x = exp, y = lnwage, colour = group)) +
  geom_smooth(data = subset(wage_data, sex == 1),
              method = "lm", se = FALSE, linewidth = 0.75) +
  geom_smooth(data = subset(wage_data, sex == 2),
              method = "lm", se = FALSE, linewidth = 1.5) +
  scale_colour_manual(values = c(Male = "blue", Female = "red")) +
  scale_y_continuous(breaks = 2:6) +
  coord_cartesian(ylim = c(2, 6)) +
  labs(x = "Potential Experience (=Age-(Years of Education +6))",
       y = "lnwage", colour = NULL) +
  theme_minimal() +/
  theme(legend.position = "inside",
        legend.position.inside = c(0.98, 0.02),
        legend.justification = c(1, 0))
```

:::

## $F$-test

The $F$-test can be used to test differences in a model across groups.[^5]

[^5]: We will look at the case of two groups, but the approach can be extended to multiple groups. 

- For example, whether the Mincer equation (from [Tutorial 1](tutorial-1.qmd)) differs by gender. 

  $$
    \ln(wage_i) = \beta_0 + \beta_1 edu_i + \beta_2 exp_i  + u_i
  $$

- We estimated this model for the male-only sample

- The quadratic term in $exp$ has been dropped for simplification

- The model has $k+1=3$ parameters

## Test:

We may wish to ask whether:

- the model for men (group = 1; e.g. "Male")

  $$
    \ln(wage_i) = \textcolor{blue}{\beta_0} + \textcolor{blue}{\beta_1} edu_i + \textcolor{blue}{\beta_2} exp_i + u_i
  $$

is the same as, 

- the model for women (group = 2; e.g. "Female")?

  $$
    \ln(wage_i) = \textcolor{red}{\gamma_0} + \textcolor{red}{\gamma_1} edu_i + \textcolor{red}{\gamma_2} exp_i + u_i
  $$

## Null hypothesis

This question implies a null hypothesis:

$$
  H_0: \textcolor{blue}{\beta_0} = \textcolor{red}{\gamma_0},\;\textcolor{blue}{\beta_1} = \textcolor{red}{\gamma_1},\;\textcolor{blue}{\beta_2} = \textcolor{red}{\gamma_2}
$$

against, 

$$
  H_1: H_0\;\text{false}
$$

## Approach 1

The $F$-test for this hypothesis has the form:

$$
  F\text{-stat} = \frac{(SSR_P - (\textcolor{blue}{SSR_1} + \textcolor{red}{SSR_2}))/(k+1)}{(\textcolor{blue}{SSR_1} + \textcolor{red}{SSR_2})/(n-2(k+1))}
$$

where, 

- $SSR_P$ is the "pooled" ($P$) $SSR$ from a regression of the joint sample. 
- $\textcolor{blue}{SSR_1}$ is the $SSR$ from a regression for just group 1.
- $\textcolor{red}{SSR_2}$ is the $SSR$ from a regression for just group 2.

  - together $\textcolor{blue}{SSR_1}$ and $\textcolor{red}{SSR_2}$ represent the unrestricted model. 

- $n-2(k+1)$ is the **combined** degrees of freedom from group 1 and 2. 

  - $k+1$ parameters are estimated in each group. 
  
## Distribution

The $F$-stat is (unsurprisingly) $F$-distributed,

$$
  F\text{-stat} \underset{H_0}{\sim} F_{k+1,n-2(k+1)}
$$

## Approach 2

This test has an equivalent form using interactions. Let, 

$$
  d_i= \mathbf{1}\{\text{`Group 2'}\}
$$

For example, 

$$
  d_i= \mathbf{1}\{\text{`Female'}\}
$$

## Interactions

We can write the unrestricted model as, 

$$
  \ln(wage_i) = \beta_0 + \beta_1 edu_i+\beta_2 exp_i +\beta_3 d_i + \beta_4 edu_i\times d_i + \beta_5 exp_i\times d_i +u_i
$$

We now have 6 parameters: $2\times (k+1)$ (for $k=2$)

##

As before, the dummy and dummy-interactions tell you about differences,

$$
  \beta_3 = E[\ln(wage_i)|\textcolor{red}{d_i=1},edu_i=0,exp_i=0]-E[\ln(wage_i)|\textcolor{blue}{d_i=0},edu_i=0,exp_i=0]
$$

and

$$
  \beta_j = \frac{\partial E[\ln(wage_i)|\textcolor{red}{d_i=1},edu_i,exp_i]}{\partial x_{ij}}-\frac{\partial E[\ln(wage_i)|\textcolor{blue}{d_i=0},edu_i,exp_i]}{\partial x_{ij}} \qquad j=4,5
$$
  
## Test

Intuitively, the test for whether the model differs by group is given by,

$$
  H_0: \beta_3 = 0,\;\beta_4=0,\;\beta_5=0
$$

where,

- **unrestricted** model is given by,

$$
  \ln(wage_i) = \beta_0 + \beta_1 edu_i+\beta_2 exp_i +\beta_3 d_i + \beta_4 edu_i\times d_i + \beta_5 exp_i\times d_i +u_i
$$

  - $df_{ur} = n-6$

##

- **restricted** model is given by,

$$
  \ln(wage_i) = \beta_0 + \beta_1 edu_i + \beta_2 exp_i +v_i
$$

  - $df_{r} = n-3$
  
## $F$-stat

The $F$-stat is given by, 

$$
  F\text{-stat} = \frac{(SSR_{r}-SSR_{ur})/3}{SSR_{ur}/(n-6)} 
$$

::: {.fragment .fade-in}
Compare this to the first approach:

$$
  F\text{-stat} = \frac{(SSR_P - (\textcolor{blue}{SSR_1} + \textcolor{red}{SSR_2}))/3}{(\textcolor{blue}{SSR_1} + \textcolor{red}{SSR_2})/(n-6)}
$$

:::

## Equivalence

These two test-statics are equivalent because, 

1. $SSR_P=SSR_r$

2. $\textcolor{blue}{SSR_1} +\textcolor{red}{SSR_2} = SSR_{ur}$

## Demonstration {.scrollable}

Here are the SSR for each version:

::: {.panel-tabset}

### Stata

- Model for group 1 (i.e. "male")

```{stata}
*| echo: true
qui reg lnwage edu exp if sex==1
dis e(rss)
```

- Model for group 2 (i.e. "female")

```{stata}
*| echo: true
qui reg lnwage edu exp if sex==2
dis e(rss)
```

- Unrestricted model (i.e. with interactions)

```{stata}
*| echo: true
qui reg lnwage i.sex##c.edu i.sex##c.exp
dis e(rss)
```

- Restricted model (i.e. without interactions)

```{stata}
*| echo: true
qui reg lnwage edu exp
dis e(rss)
```

### R

```{r}
#| code-fold: true

# make sex a factor variable
wage_data$sex <- factor(wage_data$sex)
```

- Model for group 1 (i.e. "male")

```{r}
#| echo: true
model_m <- lm(lnwage ~ edu + exp, data = wage_data, subset = sex == 1)
deviance(model_m)
```

- Model for group 2 (i.e. "female")

```{r}
#| echo: true
model_f <- lm(lnwage ~ edu + exp, data = wage_data, subset = sex == 2)
deviance(model_f)
```

- Unrestricted model (i.e. with interactions)

```{r}
#| echo: true
model_ur <- lm(lnwage ~ sex * edu + sex * exp, data = wage_data)
deviance(model_ur)
```

- Restricted model (i.e. without interactions)

```{r}
#| echo: true
model_r <- lm(lnwage ~ edu + exp, data = wage_data)
deviance(model_r)
```

:::

## {.scrollable}

Here is how you would programme the test

::: {.panel-tabset}

### Stata

- Estimate interacted model

```{stata}
*| echo: true
qui reg lnwage i.sex##c.edu i.sex##c.exp
test 2.sex 2.sex#c.edu 2.sex#c.exp
```

The notation `2.sex` is how Stata names the coefficient on level `2` of the factor variable `sex`, and `2.sex#c.edu` is the interaction with `c.edu` (a `c`ontinuous variable). 

### R

- Estimate interacted model

```{r}
#| echo: true
model_ur <- lm(lnwage ~ sex * edu + sex * exp, data = wage_data)
linearHypothesis(model_ur, c("sex2 = 0", "sex2:edu = 0", "sex2:exp = 0"))
```

The notation `sex2` is how R names the coefficient on level `2` of the factor `sex`, and `sex2:edu` its interaction with `edu` (a continuous variable).
:::



# References

## Bibliography
