---
title: Basic Programming in R
format: 
  html: default
engine: knitr  
---

The goal for this short handout is to introduce you to some basic programming tools within R. [Lab 1](lab-1-R.qmd) focuses on the tools needed to open, manipulate, and summarize data. Here you will learn about some important programming features:

- how to create objects: scalars, sequences, vectors, lists
- how to automate repetitive processes using loops

## Single values

We begin by assign an object `x` a value using the operator `<-`. 

```{r}
x <- 3 + sin(pi/2)
```

If you're working in RStudio, `x` (=4) will appear as a stored value under the "Environment" tab. We can now use `x` in various computations; for example, solving for its square root.

```{r}
sqrt(x)
```

Note, doing so does not change the stored value of `x`. We could define a new value `y` as its square root or replace `x` with its square root. 
```{}
y <- sqrt(x)
x <- sqrt(x)
```

The store value of `x` (and `y`) is now 2. 

Values need not be numerical. They can also stored characters (often referred to as strings in other languages/software). For example, 

```{r}
z <- "United Kingdom"
```

## Sequences and Loops

Sequences are used extensively in programming. The most common place is in a loop that iterates through a sequence. We can define a sequence of numbers on in two ways. Consider the sequence $1,\dots,10$. 

```{r}
u <- seq(1,10)
v <- 1:10
```

Both give the same result, but the `seq()`-uence function is more flexible since it allows you adapt the interval. For example, we can we can make the intervals in the sequence $0.5$. 

```{r}
U <- seq(1,10,0.5)
```

An immediate application of a sequence is in a loop. Consider a loop over the sequence $1(1)10$ that simply prints the number of the loop. 

```{r}
for(i in 1:10){
  print(i)
}
```

## Vectors and Matrices

When stored as values, sequences are essentially vectors. A more flexible function for creating a vector is the `c()`-ombine function. We can combine a set of objects, numeric or character. 

```{r}
w <- c(12,1,2,3)
W <- c("12","1","2","3")
```
Vectors have two important features: length and type. 
```{r}
length(w)
typeof(w)
typeof(W)
```
You can even include a sequence within the combination. 

```{r}
w <- c(12,1:3)
w
```

When you print `w` (as above), it does so as a single row. However, if you use the `t()`-ranspose function, you will see that it displays as a row vector. Taking the transpose of the transpose shows you that the default is in fact to think about `w` as a column vector. 

```{r}
t(w)
t(t(w))
```
Here are a few fuctions that may be useful:
```{r}
# minimum
min(w)
# maximum
max(w)
# sort
sort(w)
```
Note, `sort(w)` outputs a sorted version of `w`, but does not change the stored order of values in `w`. 
Multiplication/division by a scalar, is element by element. 
```{r}
w*2
w/2
```
We can combine two columns to form a matrix. 
```{r}
A <- cbind(w,seq(1,4))
A
B <- rbind(w,seq(1,4))
B
```
Interestingly, this shows us that R does not fix the row-column dimension of a vector. Else, `rbind()` would give you a $8\times 1$ vector. We can achieve this result by first fixing `w` as a $4\times 1$ matrix. 
```{r}
C <- rbind(matrix(w),matrix(seq(1,4)))
C
```
When working with vectors (and matrices) it is important to know how to index specific values. For example, if you want the third value of `w` you can say,
```{r}
w[3]
```

## Lists

Lists can combine multiple types of data and/or values. They can become relatively complex collections, that employ a hierarchical structure to navigate. When you scrape data from a website it will usually be structured as a list. 

```{r}
site <- list("22 January 2025","United Kingdom",1:10,A)
typeof(site)
length(site)
```
The above list has length 4 as it includes for items. The third item is the sequence $1(1)10$, which has 10 elements. We can index a particular element of the list by indexing. 
```{r}
site[3]
typeof(site[[3]])
length(site[[3]])
```
Lists can be recursive
```{r}
page <- list(list("22 January 2025","United Kingdom"),list("URL","download-date"))
```
You can then extract a sublist,
```{r}
page1 <- page[[1]]
element11 <- page1[[1]]
```

## Loops

There are three main types of loops in R: `for`, `while`, and `repeat`. Here we will cover the first two.

`for` loops are probably the most common. They allow you to iterate over an argument that follows a sequence:

```{r}
for (i in 1:5) {
  print(i)
}
```

This may seem simple, but consider the fact that this sequence could be used as an index. Earlier we defined the vector `W`. 

```{r}
n <- length(w)
for(i in 1:n){
  print(W[i])
}
```

Here, I use the `length()` function to know the stopping point of the loop. 

The sequence need not increase by increments of 1 each time; for example, 

```{r}
for (i in seq(0, 20, by = 5)) {
  print(i)
}
```

`while` loops require you to set a break point: the value of the argument at which the loop stops. You also need to set the starting point. 

```{r}
i <- 1
while (i <= 5) {
  print(i)
  i <- i + 1
}
```

This loop does the same thing as the first `for` loop. 